---
title: "Chapter 2 Notes and Exercises"
output:
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---
## First Steps
```{r}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```
Research Question: Do penguins with long flippers weigh more or less than panguins with shorter flippers?

## The `penguins` data frame
-   *data frame*: a rectangular collection of variables (columns) and observations (rows)
-   *variable*: measurable quantity,quality, or property
-   *value*: the measurement reported for a variable at a given point of observations
-   *observation*: measurement made under similar conditions (a collection of variables measured)
-   *Tabular data*: set of values associated with a variable and an observation
-   *tibble*: tidyverse variant of data frames
```{r penguins-frame}
penguins # view data
glimpse(penguins) # view variable headings and first few observations
View(penguins) # view entire table
?penguins # documentation for penguins table
```

## Creating a ggplot
First, we create a frame that will house the context for our plot
```{r first-step}
ggplot(data = penguins) # nothing designated to plot; just tells what dataset will be referenced
```
Then, we'll define the relations we're looking at by determining which variables will be mapped to which axes
```{r second-step}
ggplot(
  data = penguins,
  mapping  = aes(x = flipper_length_mm, y = body_mass_g)
)# now we've added a mapping defined by the aes() function which tells us which
# variables will go with which axes

```
We can start building the visualization by mapping coordinate pairs of measurements on our axes
```{r third-step}
ggplot(data = penguins,
       mapping = aes(x=flipper_length_mm, y = body_mass_g)
       ) +
  geom_point() 
# we have told it to plot the data as points with coordinates of
# (flipper_length_mm, body_mass_g)
```
Let's add some more detail - a third variable mapped on as colors, to see if a relationship can be observed based on penguin species.
```{r fourth-step}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()
```
We can now start developing the trend line(s) to make this easier to interpret
```{r fifth-step}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```
This plot has a line for each species; if we want a single line, we need to move the species variable to a local assignment on the points, rather than a global assignment in the aesthetics.
```{r semifinal-plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")
```
to get the final state, we need shapes for each group
```{r semifinal-plot-redux}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```
Let's fix the labelling with lab() and fix it up for colorblind viewers
## Final Goal
```{r final-plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species", shape = "Species"
    ) +
  scale_color_colorblind()
```



### Exercises
#### 1 How many rows are in penguins? How many columns?
```{r solution-1}
paste("Number of rows:",
      nrow(penguins),
      "Number of columns:",
      ncol(penguins), 
      sep = "\n") |>
  cat()
```
#### 2 What does the `bill_depth_mm` variable in `penguins` data frame describe? Read the help for `?penguins` to find out.
The help documentation isn't precisely clear, but I would imagine that, where bill length is the measurement of the bill from tip to head, depth is the measure of the widest point along the beak from top beak to bottom beak.

#### 3 Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm`. That is, make a scatterplot with `bill_depth_mm` on the y-axis and `bill_length_mm` on the x-axis. Describe the relationship between these two variables.

Without accounting for species, the data seems to have little/no correlation. However, there appear to be clusters, which, upon mapping species to color, demonstrates clusters in values based around species.

```{r bill-depth-length-plot}
ggplot(penguins,
       mapping = aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(mapping = aes(color = species))
```
#### 4 What happens if you make a scatterplot of `species` vs. `bill_depth_mm`? What might be a better choice of geom?

I went with a boxplot because it felt more intuitive relative to the variables chosen.
```{r bill-depth-species-plot}
ggplot(data = penguins,
       mapping = aes(x = species, y = bill_depth_mm)) +
  geom_boxplot()
```
#### 5 Why does the code give an error, and how would you fix it?
the code `ggplot(data = penguins) + geom_point()` doesn't have mappings assigned.
I would assign two variables to this position.

#### 6 What does the `na.rm` argument do in `geom_point`? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to true.
The argument defaults to FALSE, which print a warning when missing values are identified and removed.
TRUE suppresses the warning.
```{r plot-narm}
ggplot(penguins,
       mapping = aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(mapping = aes(color = species), na.rm = TRUE)
```
#### 7 Add the following caption to the plot you made in the previous exercise.
```{r captioned-plot}
ggplot(penguins,
       mapping = aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(mapping = aes(color = species), na.rm = TRUE) +
  labs(caption = "Data come from the palmerpenguins package.")
```
#### 8 Recreate the visualization. What aesthetic should `bill_depth_mm` be mapped to? And should it be mapped at the global level or the geom level?

```{r eight-plot}
ggplot(penguins,
       mapping = aes(y = body_mass_g, x = flipper_length_mm)) +
  geom_point(mapping = aes(color = bill_depth_mm), na.rm = TRUE) +
  geom_smooth()
```

#### 9 Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.

It produced pretty much what I expected, except that I forgot the color = island is at the global level, and so would plot a new trend line for each group/island
```{r nine-plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```
#### 10 Will these two graphs look different? Why/Why not?
```{r}
x <-ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()
```

```{r}
y <- ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```
The two tables will not look different, as they are (effectively) different implementations of the same instructions.

The first passes the mappings globally, and then calls the two geometries using the mappings
The second applies the mappings locally, but they are the same mappings for each geometry, so they are functionally the same.

Given this, the former would probably be "better" code, as it is more concise, and is likely also more readable.

## ggplot2 calls

- Pro-tip from authors: " Typically, the first one or two arguments to a function are so important that you should know them by heart."
  - for ggplot this is `data` and `mapping`
  - remembering this allows us to be more concise in our code
```{r concise-plot}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(na.rm = TRUE)
```

## Visualizing Distributions
- The visualization you use depends on the variable type you're displaying
  - Categorical Variables:
    -   takes only one value from a (relatively) small set of discrete values
    -   often displayed with barplots
```{r barplot}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```
      -   when the variables have non-orered levels, arrange the bars based on their frequencies by tansforming the variable to a factor and reordering the levels
```{r freq-barplot}
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```
  - Numerical variables:
    -   take a wide range of numerical values which can be added, subtracted, or averaged while retaining meaning/coherence
    - may be discrete or continuous
    - there are a wide variety of displays; the most common univariate display is the histogram
```{r histogram}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200, na.rm = TRUE)
```
    - a density plot is also a sensisble univariate visualization
```{r density-plot}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```
### Exercises
#### 1 Make a bar plot of `species` of `penguins`, where you assign `species` to the `y` aesthetic. How is this plot different?
```{r horizontal-bar}
ggplot(penguins, aes(y = species)) +
  geom_bar()
```
It gives you a horizontal bar plot rather than the traditional vertical bars


#### 2 How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?
```{r bar-fill-border}
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```
One has a red border, the other has a red fill; the fill is more useful in most cases.
#### 3 What does the bins argument in geom_histogram() do?
The binwidth argument determines how many observations are to be included in each bin; contrasted from the `bins` argument, which determines how many bins there should be, and evenly distributes observations between them.

#### 4 Make a histogram of the `carat` variable in the `diamonds` dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?
```{r diamonds}
ggplot(diamonds, aes(x = carat)) +
  geom_density() 
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.065)
```

## Visualizing relationships
- relationship requires two (or more) variables
- comparing a numerical and a categorical variable often works by using side-by-side box plots.
```{r side-by-side}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```
 - an alternative way to compare is using overlapping density plots
```{r overlapping-density}
ggplot(penguins, aes(x = body_mass_g, color = species)) +
  geom_density(linewidth = 0.75)
```
- in the following plot, we **set** the alpha to 0.5 (half transparent)
```{r overlapping-density-alpha}
ggplot(penguins, aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)
```

- terminology:
  -   **mapping** to aesthetic: visual attribute varies with value of variable
  -   **set** the value of the aesthetic: fixed, uniform aesthetic value.

- Two categorical variables can be compared with stacked bar plots
  - First plot shows frequencies on each island; it shows a count, but not a sense of percentage
```{r stacked-bar}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```
  - second plot shows relative frequency
```{r relative-stacked-bar}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
```
- two numerical variables are most commonly visualized using the scatter plot
```{r back-to-scatter}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```
- Three or more variables can be added by adding more aesthetics or using facets
```{r four-variables}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island))
# This plot has four variables, but the more variables mapped, the harder it gets to interpret
```
```{r four-variables-faceted}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species)) +
  facet_wrap(~island)
```
### Exercises
#### 1 The `mpg` data frame that is bundled with the ggplot2 package contains 234 observations collected by the US Environmental Protection Agency on 38 car models. Which variables in `mpg` are categorical? Which variables are numerical? (Hint: Type `?mpg` to read the documentation for the dataset.) How can you see this information when you run `mpg`?
```{r}
glimpse(mpg)
?mpg
```
Manufacturer, model, transmission, drive, fuel type, and class are all categorical.
year, cyl, cty, hwy, and displ are numeric.
I found these by using glimpse() which labels the variable's data type.
#### 2 Make a scatterplot of `hwy` vs. `displ` using the `mpg` data frame. Next, map a third, numerical variable to `color`, then `size`, then both `color` and `size`, then `shape`. How do these aesthetics behave differently for categorical vs. numerical variables?
When these aesthetics are applied to a numeric, they'll scale in in a manner where their color is a function of a change in quantity, whereas if they are mapped to a categorical variable, their assignment is meaningless aside from designating a particular value/class.

```{r mpg-plot}
ggplot(mpg, aes(x = hwy, y = displ, size = displ)) +
  geom_point()
```

#### 3 In the scatterplot of `hwy` vs. `displ`, what happens if you map a third variable to `linewidth`?
```{r}
ggplot(mpg, aes(x = hwy, y = displ, linewidth = cyl)) +
  geom_point()
```
Nothing appears to happen, as points don't have a line
#### 4 What happens if you map the same variable to multiple aesthetics?
It works, but creates redundancy by communicating the same information twice (i.e. across two different aesthetics); in some cases it might work for emphasis, but more likely it will create visual clutter and either distort or otherwise inhibit interpretation.

#### 5 Make a scatterplot of `bill_depth_mm` vs. `bill_length_mm` and color the points by `species`. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by species?
Coloring by species will make the clusters more apparent.
Faceting by species also makes the clusters apparent, but doesn't show the areas of overlap; this can make interpreting by species much easier.

```{r}
ggplot(penguins, aes(x = bill_depth_mm, y = bill_length_mm)) +
  geom_point() +
  facet_wrap(~species)
```


#### 6 Why does the following yield two separate legends? How would you fix it to combine the two legends?
by using the function`labs(color = "Species")`, we set it up to create a legend for colors separately titled from the shapes. We can either remove the labs function, or add a `shape = "Species"` argument to the labs call
```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species", shape = "Species")
```
#### 7 Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?
The first one lets us know what species make up which portion of the population on each island.
The second one shows us which island each of the species is present on, and how much of their population is represented in the given island's subpopulation.
```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```


## Saving your plots
- save plots with `ggsave()`, which saves the most recently created plot to disk
- specify `width` and `height` to have control over dimensions of your plot
### Exercise
#### 1 Run the following lines of code. Which of the two plots is saved as `mpg-plot.png`? Why?
```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot.png")
```
The second one will be saved, as it is the most recently plotted, and `ggsave()` saves the most recent one.

#### 2 What do you need to change in the code above to save the plot as a PDF instead of a PNG? How would you find out what types of image files would work in `ggsave()`?
You would change the 'png' argument to 'pdf', as detailed in `?ggsave()`.
