---
title: "CH3 Exercises"
author: "Darrell A Carvalho"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# Exercises
## 1 Why does this code not work?
```{r}
my_variable <- 10
my_varıable
```
The code is not functional because the second line doesn't use i, it uses ı

## 2 Tweak each of the following R commands so that they run correctly:
```{r}
library(tidyverse) # fix spelling to library(tidyverse)

ggplot(data = mpg, aes(x = displ, y = hwy)) + # fix spelling to data
  geom_point() + # fix spelling to mapping, add comma
  geom_smooth(method = "lm") # fix syntax
```
## 3 Press Alt + Shift + K. What happens? How can you get to the same place using the menus?
It opens the shortcut cheat sheet, also available from Help>Keyboard Shortcuts Help

## 4 Let's revise an exercise from Section 2.6. Run the following lines of code. Which of the two plots saved as `mpg-plot.png`? Why?
```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```
my_bar_plot will be saved, because it is assigned to a variable and that variable is called in the function; it is only when no variable is called with a plot assigned that ggsave sill save the most recently created plot.
