---
title: "Chapter 4 Notes and Exercises"
output:
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---
## First Steps
```{r prerequisites}
library(nycflights13)
library(tidyverse)
```


## Introduction
- This chapter focuses on data transformation, often with *dplyr*
- Focuses on working on rows and columns of data frames, followed by instructions on use of the pipe to combine verbs.
- Teaches how to work with groups
### nycflights13 - the dataset
- we'll be using `nycflights13::flights`
```{r}
flights
glimpse(flights) # glimpse shows all columns and their data types
```
### dplyr basics
- dplyr verbs share the common features:
  1. The first argument is always a data frame/tibble
  2. The subsequent argumentst ypically describe which columns to operate on, using the variable names
  3. The output is always a new data frame/tibble
- dplyr verbs are designed to do one thing and one thing well.
  - Complex tasks are completed by combining verbs with pipe operations.
- dplyr verbs belong to four groups, based on what they operate on:
  - rows
  - columns
  - groups
  - tables

## Rows
### `filter()`
- keep rows based on column values
- first argument is data frame; second and further arguments are conditions
```{r long-flight-delays}
## The following code uses filter to find flights with delays over 2 hours
flights |>
  filter(dep_delay > 120)
```
```{r new-year-flights}
flights |> 
  filter(month == 1 & day == 1)
```
```{r jan-feb-flights}
flights |> 
  filter(month == 1 | month ==2)
```
- an alternative to combining `|` and `==` (or and equals) is using `%in%`
```{r flights-in-jan-feb}
flights |> 
  filter(month %in% c(1, 2))
```
- *dplyr* verbs, including `filter()`, never modify their inputs. To preserve results, assign them to variables
```{r save-jan1}
jan1 <- flights |> 
  filter(month == 1 & day == 1)
```

### Common mistakes
- common mistakes include:
  - using `=` instead of `==` for equality tests
  - forgetting to include column names in all sides of the `|` tests

### `arrange()`
- `arrange()` sorts rows based on column values, sorting and then sub sorting by columns fed into the second and subsequent arguments.
```{r flight-sort}
flights |> 
  arrange(year, month, day, dep_time)
```
  - `desc()` can be used to sort by descending order rather than the default ascending order
```{r flight-desc}
flights |> 
  arrange(desc(dep_delay))
```

### `distinct()`
- `distinct()` finds distinct rows, and distinct combinations of rows across columns
```{r remove-duplicates}
flights |> 
  distinct()
```
```{r origin-destination-pairs}
flights |> 
  distinct(origin,dest)
```
- use `.keep_all = TRUE` to keep all columns in the result
```{r origin-destination-all-columns}
flights |> 
  distinct(origin, dest, .keep_all = TRUE)
```
- `distinct()` takes the first occurrance and discards the rest; if you want to know a count, you'll have to use `count()` instead
```{r count-demo}
flights |> 
  count(origin, dest, sort = TRUE)
```

### Exercises
#### 1 Use the text below to solve the problem
In a single pipeline, find all flights that meet each of the following conditions:

    Had an arrival delay of two or more hours
    Flew to Houston (IAH or HOU)
    Were operated by United, American, or Delta
    Departed in summer (July, August, and September)
    Arrived more than two hours late, but didn’t leave late
    Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
flights |> 
  filter(dest %in% c('IAH', 'HOU'),
         carrier %in% c('UA', 'AA', 'DL'),
         month %in% c(7, 8, 9),
         (arr_delay >= 120 & dep_delay <= 0) | 
           (dep_delay >= 60 & (dep_delay - arr_delay > 30))
         )
```

#### 2 Sort `flights` to find the flights with longest departure delays. Find the flights that left earliest in the morning.
```{r long-delays}
flights |> 
  arrange(desc(dep_delay))
```

```{r earliest-departures}
flights |> 
  arrange(hour, minute)
```

#### 3 Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)
```{r fastest-flight}
flights |> 
  arrange(desc(distance / air_time))
```

#### 4 Was there a flight on every day of 2013?
```{r}
flights |> 
  filter(year == 2013) |> 
  distinct(month, day) |> 
  arrange(month)
```

#### 5 Which flights traveled the farthest distance? Which traveled the least distance?
```{r farthest-distance}
flights |> 
  arrange(desc(distance))
```
```{r}
flights |> 
  arrange(distance)
```

#### 6 Does it matter what order you used `filter()` and `arrange()` if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.
it is better to arrange first, and then filter afterwards, because otherwise the filter has to work through the unsorted data, which is significantly slower. in the demonstrations below, arrange -> filter is 2.98 times faster than filter -> arrange

```{r}
start.time <- Sys.time()

flights |> 
  filter(distance >=30) |> 
  arrange(desc(arr_time))
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

```

```{r}
start.time <- Sys.time()

flights |> 
  arrange(desc(arr_time)) |> 
  filter(distance >=30)
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

## Columns
### `mutate()`
- creates new columns from existing columns
```{r gain-and-speed}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```
- by default, new columns append to the right
  - you can place them at different locations using the argument `.before` or `.after`in the mutate call
- you can choose to only keep certain columns on creation by using the `.keep` argument
  - `.keep = "used"` keeps only those columns used in, or created by, the mutate function call
```{r used}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```
  
### `select()`
- changes which columns are present, retaining only the selected
  - select columns by name:
```{r name-select}
flights |> 
  select(year, month, day)
```
  - select between two columns (inclusive):
```{r select-between}
flights |> 
  select(year:day)
```
  - select all except excluded range:
```{r excluded-range}
flights |> 
  select(!year:day)
```
  - select only character columns:
```{r character-columns}
flights |> 
  select(where(is.character))
```
  - rename with select:
```{r select-rename}
flights |> 
  select(tail_num = tailnum)
```

### `rename()`
- changes column names/headers
  - to rename while keeping all rows, use `rename()` instead of `select()`
```{r rename}
flights |> 
  rename(tail_num = tailnum)
```

### `relocate()`
- changes column position
  - by default, `relocate()` moves variables to the front/left
  - you may also use `.before` and `.after` to determine placement
```{r}
flights |> 
  relocate(time_hour, air_time)
flights |> 
  relocate(year:dep_time, .after = time_hour)
flights |> 
  relocate(starts_with("arr"), .before = dep_time)
```

### Exercises
#### 1 Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?
`sched_dep_time` + `dep_delay` = `dep_time`
`dep_time` - `sched_dep_time` = `dep_delay`
`dep_time` - `dep_delay` = `sched_dep_time`

#### 2 Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.
```{r}
flights |> 
  select(ends_with("time"), 
         ends_with("delay"), 
         -starts_with("sched"), 
         -starts_with("air"))
flights |> 
  select(dep_time:arr_delay, 
         -starts_with("sched"), 
         -starts_with("air"))
```


#### 3 What happens if you specify the name of the same variable multiple times in a `select()` call?
```{r}
flights |> 
  select("year", "year", "year")
```
It only selects the column once.


#### 4 What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?
```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")

flights |> 
  select(any_of(variables))
```
`any_of()`takes all the strings passed to it and finds the columns with those names, returning all that exist without throwing errors for those missing.


#### 5 Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?
```{r}
flights |> select(contains("TIME"))
```

#### 6 Rename `air_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.
```{r}
flights |> 
  rename(air_time_min = air_time) |> 
  relocate(air_time_min)
```


#### 7 Why doesn’t the following work, and what does the error mean?
```{r}
flights |> 
  select(tailnum) |> 
  arrange(arr_delay)
```
The code doesn't work because the select drops the other columns, including arr_delay; the error means that the non-existent column cannot be found in the new, smaller data frame.

## The pipe
- pipe makes verb combination much easier:
```{r fast-flights-houston}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```
- pipes also (arguably) make code more readable


## Groups
- three key tools for working with groups:
  - `group_by()`
  - `summarize()`
  - `slice_` functions
### `group_by()`
- `group_by()` divides dataset into groups:
```{r flight-month}
flights |> 
  group_by(month)
```
- subsequent verbs function off the grouping variable
### `summarize()`
- reduces data frame to summary statistics for group(s)
```{r month_summary}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE)
  )
# note the na.rm = TRUE argument is necessary, or else NA values will prevent meaningful means from being computed, replacing them with NA
```
- multiple summaries can be created in a single `summarize()` call
```{r mean-and-count}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE),
    count = n()
  )
```

### The `slice_` functions
- five key `slice_` functions:
  1. `slice_head(n)` takes the first **n**th row(s) from each group
  2. `slice_tail(n)` takes the last **n**th row(s) from each group
  3. `slice_min(x, n)` takes the smallest **n**th row(s) in column x per group
  4. `slice_max(x, n)` takes the largest **n**th row(s) in column x per group
  5. `slice_sample(n)` takes **n** random row(s) from each group
- n can be any number to vary the number of row(s) selected
- you can also use `prop = ` to select proportions (i.e. `0.1` for 10% of rows in each group)
- `slice_min()` and `slice_max()` keep ties unless you use the argument `with_ties = FALSE`

```{r most-delayed}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |> 
  relocate(dest)
```

### Grouping by multiple variables
- Grouping can be done on more than one variable
```{r group_by_date}
daily <- flights |> 
  group_by(year, month, day)
daily
```
- when you summarize a tibble grouped by multiples, each summary peals off the last group
  - you can avoid the warning by explicitly enabling this functionality with `.groups = "drop_last"`
  - you can keep all groups by using `.groups = "keep"`
  - you can drop all groups by using `.groups = "drop"`
```{r}
daily |> 
  summarize(n = n())

daily |> 
  summarize(n = n(), .groups = "drop")

daily |> 
  summarize(n = n(), .groups = "keep")
```

### Ungrouping
- if you want to ungroup without using `summarize()`, use `ungroup()`
```{r daily-ungroup}
daily |> 
  ungroup()
```

### `.by`
- the `.by` argument is experimental per-operation grouping.
```{r}
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = month
  )
```
- you can also group by multiple variables
```{r}
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = c(origin, dest)
  )
```
- `.by` works with all verbs
  - doesn't require the use of `.groups` to suppress warnings
  - doesn't require use of `ungroup()` to ungroup when done

### Exercises
#### 1 Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights |> group_by(carrier, dest) |> summarize(n()))
```{r}
flights |> 
  group_by(carrier) |> 
  summarize(delay = mean(dep_delay, na.rm = TRUE)) |> 
  arrange(desc(delay))
```
There are a few ways to approach the problem, but given the diversity of airports (how many carriers they serve, how big the airports are) and the airlines, any analysis would be suspect without existing domain knowledge.


#### 2 Find the flights that are most delayed upon departure from each destination.
```{r}
flights |>
  group_by(origin) |> 
  slice_max(dep_delay, n = 1) |> 
  relocate(dep_delay)
```


#### 3 How do delays vary over the course of the day. Illustrate your answer with a plot.
```{r}
delays <- flights |> 
  group_by(hour) |> 
  summarize(delay = mean(dep_delay, na.rm = TRUE))

ggplot(delays, aes(x = hour, y = delay)) +
  geom_point() +
  geom_smooth(se = FALSE)
```


#### 4 What happens if you supply a negative n to slice_min() and friends?`
"A negative value of n or prop will be subtracted from the group size"


#### 5 Explain what count() does in terms of the dplyr verbs you just learned. What does the sort argument to count() do?
"count() lets you quickly count the unique values of one or more variables"
 - it acts as a shortcut function to the linked verbs 
 `group_by() |> summarize(n = n())`
 - the `sort` argument sorts the groups' values from largest to smallest if TRUE

#### 6
Suppose we have the following tiny data frame:
```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

  a . Write down what you think the output will look like, then check if you were correct, and describe what `group_by()` does.
  The output will be the same, except it will be grouped.
```{r}
df |>
  group_by(y)
```

  b. Write down what you think the output will look like, then check if you were correct, and describe what `arrange()` does. Also comment on how it’s different from the `group_by()` in part (a)?
  The output will be sorted alphabetically on y, but will not be grouped
```{r}
df |>
  arrange(y)
```

  c. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.
  The output will be a tibble grouped on y, returning the mean x for each value y;
  the pipe passes df to group-by, then passes it to summarize().
```{r}
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
```

  d. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then, comment on what the message says.
  The output will will have more rows, as it will provide the mean x for each unique combination group of y, z.
  As it does not have the `.group` argument set to "keep", it prints out a warning that the z grouping is dropped in the output.
```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

  e. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. How is the output different from the one in part (d).
  The following will have the same results as above, except all groupings will be dropped from the output.
```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

  f. Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does. How are the outputs of the two pipelines different?
  The first pipeline will give the mean x for each combo y,z.
  The second pipeline will do likewise, but add it as a column, rather than only returning the summarized column and the called values - as such, all values of x in the x column will remain visible, and there will be duplicate entries for the combo a,L accross all rows.
```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```

## Case study
The case study focuses on baseball daya from **Lahman**
- we will compare proportion of player `H`its vs their times `AB` (At Bat):
```{r}
batters <- Lahman::Batting |> 
  group_by(playerID) |> 
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
batters
```
Below we will plot the performance against at bat count; note two things:
  1. Variation in `performance` is larger with fewer opportunities at bat (law of large numbers)
  2. positive correlation between `performance` and `n` - teams put better batters at bat more.
```{r}
batters |> 
  filter(n > 100) |> 
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1 / 10) +
  geom_smooth(se = FALSE)
```
```{r}
batters |> 
  arrange(desc(performance))
```
 - note that higher performance scores in a ranking are dominated by people with fewer at-bats who happened to get hits on those at-bats; not necessarily the best players.
