---
title: "Chapter 6 Notes and Exercises"
output:
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

## Setup
```{r}
library(tidyverse)
```
## Tidy Data
- Tidy datasets follow three rules:
  1. Each variable is a column (each column a variable)
  2. Each observation is a row (each row an observation)
  3. Each value is a sell (each cell a single value)
- Tidy datasets have two major advantages:
  1. consistency in data structure
  2. placing variables in columns takes advantage of vectorization.
- Tidy data structure is the defining element of the tidyverse - all tidyverse packages, by definition, work with tidy data.

```{r demo}
# Rate per 10,000 of TB
table1 |> 
  mutate(rate = cases / population * 10000)

# Compute total cases per year
table1 |> 
  group_by(year) |> 
  summarize(total_cases = sum(cases))

# Visualize changes over time
ggplot(table1, aes(x = year, y = cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country, shape = country)) +
  scale_x_continuous(breaks = c(1999, 2000)) # x-axis breaks at 1999 and 2000
```
### Exercises
#### 1. For each of the sample tables, describe what each observation and each column represents.
In table2, each observation is a country/year/type combination, resulting in each country being repeated four times when it should only be twice (one for each year).
the type column contains two variables - the case variable, and the population variable. The count column contains the value for each variable for each observation.

This will require a pivot_wider() call as demonstrated in exercise 2, code-block 1.

In table3, each observation is a country/year pair,the year column is correct, but rate is a composite variable title masking that the cells in the rate column contain two variables - case, and population - separated by a / character.

We will need to separate on the `/` character in rate, to plug it into a case column and population column.
#### 2. Sketch out the process you’d use to calculate the rate for table2 and table3. You will need to perform four operations
```{r table2}
table2 |> 
  pivot_wider(
    names_from = "type",
    values_from = "count"
  ) |> 
  mutate(rate = cases/population * 100000)
```
```{r table3}
table3 |> 
  separate(
    col = rate,
    into = c("cases", "population")
  ) |> mutate(
    cases = parse_number(cases), 
    population = parse_number(population), 
    rate = cases/population * 100000)
```


## Lengthening Data
- Most data is untidy
  1. often, it is structure to make data entry easier
  2. Tidy data principles are not a widely implemented set of basic principles

- Often, the first step to tidying is **pivot**ing the data.
- `tidyr` provides two **pivot** functions:
  1. `pivot_longer()`
  2. `pivot_wider()`

### Data in column names
The `billboard` dataset contains billboard hits from 2000:
```{r billboard}
billboard
```
- each observation is a song
  - column 1 is artist
  - column 2 is track
  - column 3 is date.entered
  - remaining columns are the song rank for each of 76 weeks
```{r tidy-billboard}
billboard_longer <- billboard |> pivot_longer(
  cols = starts_with("wk"), # column to be pivoted
  names_to = "week", # name of the variable stored in the column name
  values_to = "rank", # names the variable stored in the cell values
  values_drop_na = TRUE # drop NAs introduced by the pivot
) |> 
  mutate(
    week = parse_number(week) # convert week variable to number instead of string
  )
```
```{r}
billboard_longer |> 
  ggplot(aes(x = week, y = rank, group = track)) + 
  geom_line(alpha = 0.25) + 
  scale_y_reverse()
```

### How Pivoting Works
```{r}
df <- tribble(
  ~ id, ~bp1, ~bp2,
  "A",   100,  120,
  "B",   140,  115,
  "C",   120, 125
)
```

We want this dataset to have three variables:
 - `id` (already exists)
 - `measurement` (column names)
 - `value` (cell values)
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value")
```
  
- The column names are added to a new variable, defined by `names_to`
  - they are each repeated one time for each row in the original dataset
- The cell values are added to a new variable, defined by `values_to`
  - they are unwound row by row, looping around.
  
### Many variables in column names
Some datasets are more complicated - a column name may have more information embedded than one variable
```{r}
who2
```
- country and year are individual variables
- following variables have a format of XX_y_NNNN
  - the first portion is a variable for diagnostic method (sp, rel, or ep)
  - the second portion is a variable for sex/gender (m or f)
  - the third portion is a agerange band - 014 is 0 to 14; 1524 is 15-24
This breaks down to 6 variables measured in each observation:
  1. country
  2. year
  3. diagnostic method
  4. gender
  5. age range
  6. count of patients per variable (cell values)
```{r}
who2 |> 
  pivot_longer(
    cols = !(country:year), # everything but country and year because they're fine
    names_to = c("diagnosis", "gender", "age"),
    names_sep = "_", # in the name, each category is separated by underscore
    values_to = "count", # we're generating a count
  )
```

### Data and variable names in column headers
```{r}
household
```
household has the following variables:
  - family (observation number)
  - dob (date of birth)
  - name (child name)
  - child (first or second, 1 or 2)
the last three are all embedded in the column headings
- For this, we use `.value` in the `names_to` argument:
  - this tells `names_to` that the first component should stay as a variable/column header for its own variable
```{r}
household |> 
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"), # keeps dob variable, create child variable
    names_sep = "_",
    values_drop_na = TRUE
  )
```
## Widening Data
- `pivot_wider()` increases columns and reduces rows
  - it helps gather an observation that is spread across multiple rows
`cms_patient_experience` collects patient experience of Medicare and Medicaid
```{r}
cms_patient_experience
```

- core unit is organization; each org is spread across six rows
  - each row is a measurement taken for the org
```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```
Let's pivot:
```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"), # identifies the column containing observation ids
    names_from = measure_cd, # column(s) containing column headers to use
    values_from = prf_rate # column(s) to get cell values from
  )
```
### How does `pivot_wider()` work?

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115,
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> pivot_wider(
  names_from = measurement,
  values_from = value
)
```

Behind the scenes, it operates by finding the distinct entries to use as titles
- it pulls them as a vector to be plugged in to the later
```{r}
df |> 
  distinct(measurement) |> 
  pull()
```
It takes all columns not going into the new names or values and finds the distinct entries to determine the rows/observations:
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct()
```

It combines the two to make an empty data frame:
```{r}
df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(bp1 = NA, bp2 = NA, bp3 = NA)
```

It then fills in the cells with the missing values.
